The Numerical Accuracy of Computing Electrostatic Interaction in the Inhomogeneous and Correlated Molecular Systems: for the Ewald Summation, SPME and Staggered Mesh Ewald Methods.

### 2012-05-01 longRange.hetero/publish/sendfile/main.tex
add sendfile

\title{The Error Estimate for Computing Electrostatic Interaction
  in the Inhomogeneous and Correlated Molecular Systems:
  for the Ewald Summation, SPME and Staggered Mesh Ewald.
}
\author{Han Wang}
\email{han.wang@fu-berlin.de}
\affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}
\author{Pingwen Zhang}
\affiliation{LMAM and School of Mathematical Sciences, Peking University}
\author{Christof Sch\"utte}
\affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}

\begin{abstract}
  In this work, we develop the accurate error estimates for
  some state-of-art algorithms of long-range electrostatic interaction
  in inhomogeneous and correlated molecular systems.
  They are the Ewald summation, the smooth particle mesh Ewald (SPME) and
  the staggered mesh Ewald method. Two branches of force computation, namely
  the ik- and analytical differentiation, are considered. All the
  estimates are developed by proposing a more general framework:
  if the error force is of pairwise form,
  then the root mean square force error is composed by three additive parts,
  the homogeneity error, the inhomogeneity error and the correlation
  error. Computationally scalable estimates (estimating the errors
  at the cost $\mathcal O(N\log N)$) are developed for all the considered
  algorithms.
  The effectiveness of the proposed estimates and the important role of
  the correlation error are carefully checked and demonstrated by
  example systems.
\end{abstract}

@@

Due to symmetricity of $\v g(\v m)$ and $A(\v m, \v r)$, the

@@

\begin{figure}
  \centering
  \includegraphics[]{fig.new/rand1/fig.rho.eps}
  \caption{
    The charge number density  of Example 2.
    The red line: positive charge. The blue line: negative charge.
    Since the charges are uniformly distributed
    on $y$ and $z$ direction, the densities are averaged on $y$ and $z$,
    and plotted as a
    function of $x$.
  }
  \label{fig:tmp-rho1}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[]{fig.new/fig.rand1.error.eps}
  \caption{
    Example 1: the real RMS errors and the corresponding
    estimates. Different colors denote
    different parts of the Ewald summation. Red: the direct
    part of Ewald summation. Green: the reciprocal
    part of Ewald summation, calculated by truncation.
    ``$\times$'' denotes the real error, and the solid line denotes
    the estimated error.
    Blue: the reciprocal part of Ewald summation, calculated
    by fast algorithms. ``$\blacksquare$'': real error of
    the original analytical
    differentiation. ``$\boxdot$'': real error of the original ik-differentiation.
    ``$\times$'': real error of the staggered mesh analytical differentiation.
    ``$+$'': real error of the staggered mesh ik-differentiation.
    Dashed line: estimated original analytical differentiation.
    Dotted line: estimated original ik-differentiation.
    Solid line: estimated staggered mesh ik-/analytical differentiation.
    The RMS errors are averaged over $y$ and $z$ directions, and are
    plotted against $x$ axis.
    The cut-off in the real space is 1.31 \textsf{nm}, the number of
    freedom in the reciprocal space is $120\times 60\times 60$, the
    parameter $\beta$ is $2.5\; \textsf{nm}^{-1}$, and the order of
    B-spline interpolation is 6.
    }
    % Example 1: the actual (denoted by dots) and estimated
    % (denoted by straight lines) RMS force error of the direct part
    % $\sqrt{\langle\vert\Delta \v F_{\textrm{dir}}\vert^2\rangle}$
    % (red), reciprocal part of Ewald summation
    % $\sqrt{\langle\vert\Delta \v F_{\textrm{rec}}\vert^2\rangle}$
    % (green) and the reciprocal part of the analytical differentiation
    % $\sqrt{\langle\vert\Delta \v
    %   F^{\textrm{ana}}_{\textrm{rec}}\vert^2\rangle}$ (blue). The RMS
    % force error of analytical differentiation
    % $\sqrt{\langle\vert\Delta \v F^{\textrm{ana}}\vert^2\rangle}$ is
    % given in pink.  Since the charge distribution is uniform on the
    % $y$ and $z$ direction, the $y$ and $z$ dimension of the mean error
    % force is averaged, and the errors are plotted on $x$ direction.
    % The cut-off in the real space is 3 \textsf{nm}, the number of
    % freedom in the reciprocal space is $100\times 50\times 50$, the
    % parameter $\beta$ is $1.0\; \textsf{nm}^{-1}$ and the order of
    % B-spline interpolation used for the analytical differentiation is
    % 6.}
  \label{fig:error1}
\end{figure}

@@

\begin{figure}
  \centering
  % \includegraphics[width=.48\textwidth]{fig.new/rand2/fig.rho.eps}
  \includegraphics[]{fig.new/rand2/fig.rho.eps}
  \caption{Charge distribution of Example 2. The red line
    denotes the positive charge distribution, while the blue line
    denotes the negative charge distribution. Since all distributions
    are uniform on $y$ and $z$ direction, they are plotted as a
    function of $x$.}
  \label{fig:tmp-rho2}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[]{fig.new/fig.rand2.meanf.eps}
  \caption{
    Example 2: the real mean error forces and the corresponding
    estimates. All symbols and working parameters are the same
    as     Fig.~\ref{fig:error1}.
    The reciprocal mean error forces vanish, so the points and lines
    overlap each other. 
    The errors are averaged over $y$ and $z$ directions, and are
    plotted against $x$ axis.
    % Red: the direct
    % part of Ewald summation. Green: the reciprocal
    % part of Ewald summation, calculated by truncation.
    % ``$\times$'' denotes the real error, and the solid line denotes
    % the estimated error.
    % Blue: the reciprocal part of Ewald summation, calculated
    % by fast algorithms. ``$\blacksquare$'': real error of
    % the oringinal analytical
    % differentiation. 
    % ``$\times$'': real error of the staggered mesh analytical differentiation.
    % Dashed line: estimated oringinal analytical differentiation.
    % Solid line: estimated staggered mesh ik-/analytical differentiation.
    % The $x$ component of the
    % error forces are averged over $y$ and $z$ directions, and are
    % plotted against $x$ axis.
    % The cut-off in the real space is 1.31 \textsf{nm}, the number of
    % freedom in the reciprocal space is $120\times 60\times 60$, the
    % parameter $\beta$ is $2.5\; \textsf{nm}^{-1}$, and the order of
    % B-spline interpolation is 6.
    }
  \label{fig:meanf2}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[]{fig.new/fig.rand2.error.eps}
  \caption{
    Example 2: the real RMS error and the corresponding
    estimates.
    The symbols and working parameters used in this figure are the same as
    Fig.~\ref{fig:error1}.
    The real and estimated error of the original ik-differentiation
    are very closed to the staggered mesh Ewald results, so
    they are not shown in this figure for clarity, 
    The errors are averaged over $y$ and $z$ directions, and are
    plotted against $x$ axis.
  }
  \label{fig:error2}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[]{fig.new//fig.water.orig.error.eps}
  \caption{
    Example 3: the real RMS errors and the corresponding
    estimates of original ik- and analytical differentiation.
    Red: the direct part.
    Green: the original ik-differentiation.
    Blue: the original analytical differentiation.
    ``$\times$'': the real error.
    Dashed line: the error estimate, without the correlation error estimate.
    Solid  line: the error estimate, with the nearest neighbor
    approximation of the correlation error.
    The working parameters are the same as Example 1.
    The black arrows indicate the improvement
    of error estimate by including nearest neighbor
    approximation of correlation error.
  }   
  \label{fig:water-error0}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[]{fig.new//fig.water.ana.st.error.eps}
  \caption{
    Example 3: the real RMS errors and the corresponding
    estimates. 
    Blue ``$\times$'': the real error of the staggered mesh
    analytical differentiation.
    Blue ``$+$'': the real error of the staggered mesh
    ik-differentiation, it is overlapping with the analytical differentiation.
    Blue dashed line: the error estimate, without the correlation error.
    Blue solid line: the error estimate, with the nearest
    neighbor approximation of the correlation error.
    The RMS errors are averaged over $y$ and $z$ directions, and are
    plotted against $x$ axis.
    The working parameters are the same as Example 1.
    The black arrow indicates the improvement
    of error estimate by including nearest neighbor
    approximation of correlation error.
  }   
  \label{fig:water-error}
\end{figure}

@@

In this paper, we proposed the error estimates for some state-of-art
algorithms calculating the long-range electrostatic interaction in a
molecular system. They were the Ewald summation, the smooth particle
mesh Ewald (SPME, both ik- and analytical differentiation) and
the staggered mesh Ewald (both ik- and analytical differentiation).
Unlike the previous error estimates, the new error estimates did NOT
assume the homogeneity and uncorrelation of the system, which was a too
strong assumption for most molecular simulations of practical
interest.  A general error estimate framework was proposed
to study all the aforementioned estimates.
When the error force has a pairwise form,
the RMS error was proved to be composed by three additive parts:
the homogeneity error, the inhomogeneity error and the correlation
error.
Moreover, the framework quantitatively provided the estimate
of all these error components, and suggested a computationally
scalable way (at the cost of $\mathcal O(N\log N)$) to calculate them.

@@

The effectiveness of the proposed error estimates was verified by
three numerical tests: two ideal inhomogeneous cases, in which the charges
were uncorrelated,
and one real water system, in which the charges were correlated.
In the ideal cases, the all error estimates were sharp, even in an extreme
case: positive and negative charges were globally separated. 
In the water system,  all estimates overestimated the real error,
because of the correlation of charges.
The quality of the real space error estimates 
was acceptable.
% However, the error estimate neglecting the charge correlation
% is not applicable  for the staggered mesh Ewald method.
The error estimates for the reciprocal space fast algorithms
were greatly improved by including the first neighbor approximation
to the correlation error.
This indicated among all possible correlations,
the rigid bond and angle correlation within one molecule
was dominant.

>>>>

\title{The Numerical Accuracy of Computing Electrostatic Interaction
  in the Inhomogeneous and Correlated Molecular Systems:
  for the Ewald Summation, SPME and Staggered Mesh Ewald Methods.
}
\author{Han Wang}
\email{han.wang@fu-berlin.de}
\affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}
\author{Pingwen Zhang}
\affiliation{LMAM and School of Mathematical Sciences, Peking University}
\author{Christof Sch\"utte}
\affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}

\begin{abstract}
  In this work, we develop the accurate error estimates for
  three state-of-art algorithms of long-range electrostatic interaction
  in inhomogeneous and correlated molecular systems.
  They are the Ewald summation, the smooth particle mesh Ewald (SPME) and
  the staggered mesh Ewald methods. Two branches of force computation, namely
  the ik- and analytical differentiation, are considered. All the
  estimates are developed by proposing a more general framework:
  if the error force is of pairwise form,
  then the root mean square force error is composed by three additive parts,
  the homogeneity error, the inhomogeneity error and the correlation
  error. Computationally scalable estimates (estimating the errors
  at the cost $\mathcal O(N\log N)$) are developed for all the considered
  algorithms.
  The effectiveness of the proposed estimates and the important role of
  the correlation error are carefully checked and demonstrated by
  example systems.
\end{abstract}

@@

Due to the symmetry of $\v g(\v m)$ and $A(\v m, \v r)$, the

@@

\begin{figure}
  \centering
  \includegraphics[]{rand1.rho.eps}
  \caption{
    The charge number density  of Example 2.
    The red line: positive charge. The blue line: negative charge.
    Since the charges are uniformly distributed
    on $y$ and $z$ direction, the densities are averaged on $y$ and $z$,
    and plotted as a
    function of $x$.
  }
  \label{fig:tmp-rho1}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[]{fig.rand1.error.eps}
  \caption{
    Example 1: the real RMS errors and the corresponding
    estimates. Different colors denote
    different parts of the Ewald summation. Red: the direct
    part of Ewald summation. Green: the reciprocal
    part of Ewald summation, calculated by truncation.
    ``$\times$'' denotes the real error, and the solid line denotes
    the estimated error.
    Blue: the reciprocal part of Ewald summation, calculated
    by fast algorithms. ``$\blacksquare$'': real error of
    the original analytical
    differentiation. ``$\boxdot$'': real error of the original ik-differentiation.
    ``$\times$'': real error of the staggered mesh analytical differentiation.
    ``$+$'': real error of the staggered mesh ik-differentiation.
    Dashed line: estimated original analytical differentiation.
    Dotted line: estimated original ik-differentiation.
    Solid line: estimated staggered mesh ik-/analytical differentiation.
    The RMS errors are averaged over $y$ and $z$ directions, and are
    plotted against $x$ axis.
    The cut-off in the real space is 1.31 \textsf{nm}, the number of
    freedom in the reciprocal space is $120\times 60\times 60$, the
    parameter $\beta$ is $2.5\; \textsf{nm}^{-1}$, and the order of
    B-spline interpolation is 6.
    }
    % Example 1: the actual (denoted by dots) and estimated
    % (denoted by straight lines) RMS force error of the direct part
    % $\sqrt{\langle\vert\Delta \v F_{\textrm{dir}}\vert^2\rangle}$
    % (red), reciprocal part of Ewald summation
    % $\sqrt{\langle\vert\Delta \v F_{\textrm{rec}}\vert^2\rangle}$
    % (green) and the reciprocal part of the analytical differentiation
    % $\sqrt{\langle\vert\Delta \v
    %   F^{\textrm{ana}}_{\textrm{rec}}\vert^2\rangle}$ (blue). The RMS
    % force error of analytical differentiation
    % $\sqrt{\langle\vert\Delta \v F^{\textrm{ana}}\vert^2\rangle}$ is
    % given in pink.  Since the charge distribution is uniform on the
    % $y$ and $z$ direction, the $y$ and $z$ dimension of the mean error
    % force is averaged, and the errors are plotted on $x$ direction.
    % The cut-off in the real space is 3 \textsf{nm}, the number of
    % freedom in the reciprocal space is $100\times 50\times 50$, the
    % parameter $\beta$ is $1.0\; \textsf{nm}^{-1}$ and the order of
    % B-spline interpolation used for the analytical differentiation is
    % 6.}
  \label{fig:error1}
\end{figure}

@@

\begin{figure}
  \centering
  % \includegraphics[width=.48\textwidth]{fig.new/rand2/fig.rho.eps}
  \includegraphics[]{rand2.rho.eps}
  \caption{Charge distribution of Example 2. The red line
    denotes the positive charge distribution, while the blue line
    denotes the negative charge distribution. Since all distributions
    are uniform on $y$ and $z$ direction, they are plotted as a
    function of $x$.}
  \label{fig:tmp-rho2}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[]{fig.rand2.meanf.eps}
  \caption{
    Example 2: the real mean error forces and the corresponding
    estimates. All symbols and working parameters are the same
    as     Fig.~\ref{fig:error1}.
    The reciprocal mean error forces vanish, so the points and lines
    overlap each other. 
    The errors are averaged over $y$ and $z$ directions, and are
    plotted against $x$ axis.
    % Red: the direct
    % part of Ewald summation. Green: the reciprocal
    % part of Ewald summation, calculated by truncation.
    % ``$\times$'' denotes the real error, and the solid line denotes
    % the estimated error.
    % Blue: the reciprocal part of Ewald summation, calculated
    % by fast algorithms. ``$\blacksquare$'': real error of
    % the oringinal analytical
    % differentiation. 
    % ``$\times$'': real error of the staggered mesh analytical differentiation.
    % Dashed line: estimated oringinal analytical differentiation.
    % Solid line: estimated staggered mesh ik-/analytical differentiation.
    % The $x$ component of the
    % error forces are averged over $y$ and $z$ directions, and are
    % plotted against $x$ axis.
    % The cut-off in the real space is 1.31 \textsf{nm}, the number of
    % freedom in the reciprocal space is $120\times 60\times 60$, the
    % parameter $\beta$ is $2.5\; \textsf{nm}^{-1}$, and the order of
    % B-spline interpolation is 6.
    }
  \label{fig:meanf2}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[]{fig.rand2.error.eps}
  \caption{
    Example 2: the real RMS error and the corresponding
    estimates.
    The symbols and working parameters used in this figure are the same as
    Fig.~\ref{fig:error1}.
    The real and estimated error of the original ik-differentiation
    are very closed to the staggered mesh Ewald results, so
    they are not shown in this figure for clarity, 
    The errors are averaged over $y$ and $z$ directions, and are
    plotted against $x$ axis.
  }
  \label{fig:error2}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[]{fig.water.orig.error.eps}
  \caption{
    Example 3: the real RMS errors and the corresponding
    estimates of original ik- and analytical differentiation.
    Red: the direct part.
    Green: the original ik-differentiation.
    Blue: the original analytical differentiation.
    ``$\times$'': the real error.
    Dashed line: the error estimate, without the correlation error estimate.
    Solid  line: the error estimate, with the nearest neighbor
    approximation of the correlation error.
    The working parameters are the same as Example 1.
    The black arrows indicate the improvement
    of error estimate by including nearest neighbor
    approximation of correlation error.
  }   
  \label{fig:water-error0}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[]{fig.water.ana.st.error.eps}
  \caption{
    Example 3: the real RMS errors and the corresponding
    estimates. 
    Blue ``$\times$'': the real error of the staggered mesh
    analytical differentiation.
    Blue ``$+$'': the real error of the staggered mesh
    ik-differentiation, it is overlapping with the analytical differentiation.
    Blue dashed line: the error estimate, without the correlation error.
    Blue solid line: the error estimate, with the nearest
    neighbor approximation of the correlation error.
    The RMS errors are averaged over $y$ and $z$ directions, and are
    plotted against $x$ axis.
    The working parameters are the same as Example 1.
    The black arrow indicates the improvement
    of error estimate by including nearest neighbor
    approximation of correlation error.
  }   
  \label{fig:water-error}
\end{figure}

@@

In this paper, we proposed the error estimates for three state-of-art
algorithms calculating the long-range electrostatic interaction in a
molecular system. They were the Ewald summation, the smooth particle
mesh Ewald (SPME, both ik- and analytical differentiation) and
the staggered mesh Ewald (both ik- and analytical differentiation) methods.
Unlike the previous error estimates, the new error estimates did NOT
assume the homogeneity and uncorrelation of the system, which was a too
strong assumption for most molecular simulations of practical
interest.  A general error estimate framework was proposed
to study all the aforementioned estimates.
When the error force has a pairwise form,
the RMS error was proved to be composed by three additive parts:
the homogeneity error, the inhomogeneity error and the correlation
error.
Moreover, the framework quantitatively provided the estimate
of all these error components, and suggested a computationally
scalable way (at the cost of $\mathcal O(N\log N)$) to calculate them.

@@

The effectiveness of the proposed error estimates was verified by
three numerical tests: two ideal inhomogeneous cases, in which the charges
were uncorrelated,
and one real water system, in which the charges were correlated.
In the ideal cases, the all error estimates were sharp, even in an extreme
case: positive and negative charges were globally separated. 
In the water system,  all estimates overestimated the real error,
because of the correlation of charges.
The quality of the real space error estimates 
was acceptable.
% However, the error estimate neglecting the charge correlation
% is not applicable  for the staggered mesh Ewald method.
The error estimates for the reciprocal space fast algorithms
were impressively improved by including the first neighbor approximation
to the correlation error.
This indicated among all possible correlations,
the rigid bond and angle correlation within one molecule
was dominant.
